A comparison of design-based and model-based approaches for finite population spatial sampling and inference.

### 2022-05-19 inst/manuscript/manuscript.tex
minor updates

The GRTS algorithm selects samples by utilizing a particular mapping
between two-dimensional and one-dimensional space that preserves
proximity relationships. First, the bounding box of the domain is split
up into four distinct, equally sized squares called level-one cells.
Each level-one cell is randomly assigned a level-one address of 0, 1, 2,
or 3. The set of level-one cells is denoted by \(\mathcal{A}_1\) and
defined as \(\mathcal{A}_1 \equiv \{a_1: a_1 = 0, 1, 2, 3\}\). Within
each level-one cell, the inclusion probability for each population unit
(which is pre-specified) is summed, and if any of these sums exceed one,
a second level of cells is added. Then each level-one cell is split into
four distinct, equally sized squares called level-two cells. Each
level-two cell is randomly assigned a level-two address of 0, 1, 2, or
3. The set of level-two cells is denoted by \(\mathcal{A}_2\) and
defined as
\(\mathcal{A}_2 \equiv \{a_1a_2: a_1 = 0, 1, 2, 3; a_2 = 0, 1, 2, 3\}\).
The inclusion probabilities within each level-two cell are summed, and
if any of these sums exceed one, a third level of cells is added. This
process continues for \(k\) steps, until all level-\(k\) cells have
inclusion probability sums no larger than one. Then
\(\mathcal{A}_k \equiv \{a_1...a_k : a_1 = 0, 1, 2, 3; ...; a_k = 0, 1, 2, 3\}\).
Figure\(~\)\ref{fig:grts} provides some intuition regarding the
assignment of level-one and level-two cells.

@@

It is also important to quantify the uncertainty in \(\hat{\tau}_{ht}\).
The Horvitz-Thompson (Horvitz and Thompson, 1952) and Sen-Yates-Grundy
(Sen, 1953; Yates and Grundy, 1953) variance estimators are often used
to estimate \(\text{Var}(\hat{\tau}_{ht})\), but these estimators have
two drawbacks. First, they rely on calculating \(\pi_{ij}\), the
probability that population unit \(i\) and population unit \(j\) are
both in the sample -- this quantity can be challenging if not impossible
to calculate analytically for GRTS samples. Second, these estimators
tend to ignore the spatial locations of the population units. To address
these two drawbacks simultaneously, Stevens and Olsen (2003) proposed
the local neighborhood variance estimator. The local neighborhood
variance estimator does not rely on \(\pi_{ij}\) and estimates the
variance of \(\hat{\tau}\) conditional on the random properties of the
GRTS sample -- the idea being that this conditioning should yield a more
precise estimate of \(\hat{\tau}\). They show that the contribution from
each sampled population unit to the overall variance is dominated by
local variation. Thus the local neighborhood variance estimator is a
weighted sum of variance estimates from each sampled population unit's
local neighborhood. These local neighborhoods contain the sampled
population unit itself and its three nearest neighbors (among all other
sampled population units). For more details, see Stevens and Olsen
(2003).

@@

We evaluated performance of the four sampling-inference combinations in
36 different simulation scenarios. The 36 scenarios resulted from the
crossing of three sample sizes, two location layouts (of the population
units), two response types, and three proportions of dependent random
error (DRE). The three sample sizes (\(n\)) were \(n = 50, n = 100,\)
and \(n = 200\). Samples were always selected from a population size
(\(N\)) of \(N = 900\). The two location layouts were random and
gridded. Locations in the random layout were randomly generated inside
the unit square (\([0, 1] \times [0, 1]\)). Locations in the gridded
layout were placed on a fixed, equally spaced grid inside the unit
square. The two response types were normal and skewed. For the normal
response type, the response was simulated using mean-zero random errors
with the exponential covariance (Equation\(~\)\ref{equation:expcov}) for
three proportions of dependent random error (DRE): 0\% DRE, 50\% DRE,
and 90\% DRE. Recall the proportion of DRE is represented by
\(\sigma^2_1 / (\sigma^2_1 + \sigma^2_2)\), where \(\sigma^2_1\) and
\(\sigma^2_2\) are the DRE variance and independent random error (IRE)
variance from Equation\(~\)\ref{equation:expcov}, respectively. The
total variance, \(\sigma^2_1 + \sigma^2_2\), was always 2. The distance
parameter was always \(\sqrt{2} / 3\), chosen so that the correlation in
the DRE decayed to nearly zero at \(\sqrt{2}\), the largest possible
distance between two population units in the domain. For the skewed
response type, the response was first simulated using the same approach
as for the normal response type, except that the total variance was
0.6931 instead of 2. The response was then exponentiated, yielding a
skewed random variable whose total variance was 2. The skewed responses
were used to evaluate performance of the sampling-inference approaches
for data that were not normal but were still estimated using REML, which
relies on a normal log-likelihood. Figure \ref{fig:sim_pops} shows an
example of a realized population for the normal and skewed responses
using the random layout and 50\% DRE.

@@

The United States Environmental Protection Agency (USEPA), states, and
tribes periodically conduct National Aquatic Research Surveys (NARS) to
assess the water quality of various bodies of water in the conterminous
United States. One component of NARS is the National Lakes Assessment
(NLA), which measures various aspects of lake health and water quality.
We focus on analyzing zooplankton multi-metric indices (ZMMI) and
mercury concentrations in parts per billion (Hg ppb) from the 2012 NLA.
For ZMMI, data were collected at 1035 unique lakes. At less than 10\% of
lakes, two ZMMI replicates were collected. These were averaged for the
purposes of our study so that each lake had one measurement for ZMMI.
For Hg ppb, data were collected at 995 unique lakes (there were no
replicates). The ZMMI and Hg ppb data are shown as spatial maps and as
histograms in Figure \ref{fig:zmmi}. The ZMMI data tend to be highest
near the coasts, lowest in the Central United States, are relatively
symmetric, and have a mean of 55.05. The Hg ppb data tend to be highest
in the Northeastern United States, lowest elsewhere, are skewed, and
have a mean of 103.16 ppb. Also in Figure \ref{fig:zmmi} are separate
spatial semivariograms for ZMMI and Hg ppb. The spatial semivariogram
quantifies the halved average squared differences (semivariance) of
responses whose separation (distance) falls within a separation class.
The spatial semivariance is closely related to the spatial covariance,
and spatial semivariograms are often used to gauge the strength of
spatial dependence in data. Both ZMMI and Hg ppb seem to have moderately
strong spatial dependence (Figure \ref{fig:zmmi}), as the semivariance
increases steadily with distance (meaning that observations nearby one
another tend to be more similar than observations far apart from one
another).

\begin{figure}
\centering
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/zmmi_map.jpeg}
  \caption{Spatial map of the ZMMI population.}
  \label{fig:zmmi_map}
\end{subfigure}
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/mercury_map.jpeg}
  \caption{Spatial map of the Hg ppb population.}
  \label{fig:mercury_map}
\end{subfigure} \\
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/zmmi_hist.jpeg}
  \caption{Histogram of the ZMMI population.}
  \label{fig:zmmi_hist}
\end{subfigure}
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/mercury_hist.jpeg}
  \caption{Histogram of the Hg ppb population.}
  \label{fig:mercury_hist}
\end{subfigure} \\
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/zmmi_sv.jpeg}
  \caption{Semivariogram of the ZMMI population.}
  \label{fig:zmmi_sv_plot}
\end{subfigure}
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/mercury_sv.jpeg}
  \caption{Semivariogram of the Hg ppb population.}
  \label{fig:mercury_sv_plot}
\end{subfigure}
\caption{Exploratory graphics representing populations for the zooplankton multi-metric indices (ZMMI) and mercury concentration in parts per billion (Hg ppb) in the 2012 National Lakes Assessment (NLA) data.}
\label{fig:zmmi}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width = 1\linewidth]{figures/data_coverage.jpeg}
  \caption{NLA data interval coverage for the four sampling-inference combinations. The rows indicate the proportion of dependent error and the columns indicate the response type. The solid black lines separate the sample sizes and the dashed black lines represent 95\% coverage.}
  \label{fig:data_figconf}
\end{figure}

@@

There are several benefits and drawbacks of the design-based and
model-based approaches for finite population spatial sampling and
inference. Some we have discussed, but others we have not, and they are
worthy of consideration in future research. First, we discuss advantages
of the design-based approach. Design-based inference is often
computationally efficient, while model-based inference can be
computationally burdensome, especially for likelihood-based estimation
methods like REML that rely on the inverse of a covariance matrix.
Design-based inference easily handles binary data through a
straightforward application of the Horvitz-Thompson estimator. In
contrast, analyzing binary data using model-based inference generally
requires a logistic mixed regression model, the parameters of which can
be difficult to estimate and interpret (Bolker et al., 2009). An
advantage of design-based inference is that interval coverage is valid
(has the proper coverage rate) as long as 1) the sample is sufficiently
large to ensure the statistic's sampling distribution is approximately
normal and 2) the variance estimator is consistent (Brus and De
Gruijter, 1997; Särndal et al., 2003). This is because with the
design-based approach, the sampling plan and inclusion probabilities are
specified directly by the researcher. An advantage of SRS-DB not
previously mentioned is that it is likely to be valid given the
consistency of its variance estimator (Särndal et al., 2003). With the
model-based approach, however, interval coverage is unlikely to be valid
if the model assumptions made do not not accurately reflect reality.
Whether model assumptions accurately reflect reality can be a
challenging and sometimes impossible question to answer definitively.

>>>>

The GRTS algorithm selects samples by utilizing a particular mapping
between two-dimensional and one-dimensional space that preserves
proximity relationships. First, the bounding box of the domain is split
up into four distinct, equally sized squares called level-one cells.
Each level-one cell is randomly assigned a level-one address of 0, 1, 2,
or 3. The set of level-one cells is denoted by \(\mathcal{A}_1\) and
defined as \(\mathcal{A}_1 \equiv \{a_1: a_1 = 0, 1, 2, 3\}\). Within
each level-one cell, the inclusion probability for each population unit
(which is pre-specified) is summed, and if any of these sums are one or
greater, a second level of cells is added. Then each level-one cell is
split into four distinct, equally sized squares called level-two cells.
Each level-two cell is randomly assigned a level-two address of 0, 1, 2,
or 3. The set of level-two cells is denoted by \(\mathcal{A}_2\) and
defined as
\(\mathcal{A}_2 \equiv \{a_1a_2: a_1 = 0, 1, 2, 3; a_2 = 0, 1, 2, 3\}\).
The inclusion probabilities within each level-two cell are summed, and
if any of these sums are one or greater, a third level of cells is
added. This process continues for \(k\) steps, until all level-\(k\)
cells have inclusion probability sums less than one. Then
\(\mathcal{A}_k \equiv \{a_1...a_k : a_1 = 0, 1, 2, 3; ...; a_k = 0, 1, 2, 3\}\).
Figure\(~\)\ref{fig:grts} provides some intuition regarding the
assignment of level-one and level-two cells.

@@

It is also important to quantify the uncertainty in \(\hat{\tau}_{ht}\).
The Horvitz-Thompson (Horvitz and Thompson, 1952) and Sen-Yates-Grundy
(Sen, 1953; Yates and Grundy, 1953) variance estimators are often used
to estimate \(\text{Var}(\hat{\tau}_{ht})\), but these estimators have
two drawbacks. First, they rely on calculating \(\pi_{ij}\), the
probability that population unit \(i\) and population unit \(j\) are
both in the sample -- this quantity can be challenging if not impossible
to calculate analytically for GRTS samples. Second, these estimators
tend to ignore the spatial locations of the population units. To address
these two drawbacks simultaneously, Stevens and Olsen (2003) proposed
the local neighborhood variance estimator. The local neighborhood
variance estimator does not rely on \(\pi_{ij}\) and estimates the
variance of \(\hat{\tau}\) conditional on the random properties of the
GRTS sample -- the idea being that this conditioning should yield a more
precise estimate of \(\tau\). They show that the contribution from each
sampled population unit to the overall variance is dominated by local
variation. Thus the local neighborhood variance estimator is a weighted
sum of variance estimates from each sampled population unit's local
neighborhood. These local neighborhoods contain the sampled population
unit itself and its three nearest neighbors (among all other sampled
population units). For more details, see Stevens and Olsen (2003).

@@

We evaluated performance of the four sampling-inference combinations in
36 different simulation scenarios. The 36 scenarios resulted from the
crossing of three sample sizes, two location layouts (of the population
units), two response types, and three proportions of dependent random
error (DRE). The three sample sizes (\(n\)) were \(n = 50, n = 100,\)
and \(n = 200\). Samples were always selected from a population size
(\(N\)) of \(N = 900\). The two location layouts were random and
gridded. Locations in the random layout were randomly generated inside
the unit square (\([0, 1] \times [0, 1]\)). Locations in the gridded
layout were placed on a fixed, equally spaced grid inside the unit
square. The two response types were normal and skewed. For the normal
response type, the response was simulated using mean-zero random errors
with the exponential covariance (Equation\(~\)\ref{equation:expcov}) for
three proportions of dependent random error (DRE): 0\% DRE, 50\% DRE,
and 90\% DRE. Recall the proportion of DRE is represented by
\(\sigma^2_1 / (\sigma^2_1 + \sigma^2_2)\), where \(\sigma^2_1\) and
\(\sigma^2_2\) are the DRE variance and independent random error (IRE)
variance from Equation\(~\)\ref{equation:expcov}, respectively. The
total variance, \(\sigma^2_1 + \sigma^2_2\), was always 2. The distance
parameter was always \(\sqrt{2} / 3\), chosen so that the correlation in
the DRE decayed to nearly zero at \(\sqrt{2}\), the largest possible
distance between two population units in the domain. For the skewed
response type, the response was first simulated using the same approach
as for the normal response type, except that the total variance was
0.6931 instead of 2. The response was then exponentiated, yielding a
skewed random variable whose total variance was 2. The skewed responses
were used to evaluate performance of the sampling-inference approaches
for data that were not normally distributed but were still estimated
using REML, which relies on a normal log-likelihood. Figure
\ref{fig:sim_pops} shows an example of a realized population for the
normal and skewed responses using the random location layout and 50\%
DRE.

@@

The United States Environmental Protection Agency (USEPA), states, and
tribes periodically conduct National Aquatic Research Surveys (NARS) to
assess the water quality of various bodies of water in the conterminous
United States. One component of NARS is the National Lakes Assessment
(NLA), which measures various aspects of lake health and water quality.
We focus on analyzing zooplankton multi-metric indices (ZMMI) and
mercury concentrations in parts per billion (Hg ppb) from the 2012 NLA.
For ZMMI, data were collected at 1035 unique lakes. At less than 10\% of
lakes, two ZMMI replicates were collected. These were averaged for the
purposes of our study so that each lake had one measurement for ZMMI.
For Hg ppb, data were collected at 995 unique lakes (there were no
replicates). The ZMMI and Hg ppb data are shown as spatial maps and as
histograms in Figure \ref{fig:zmmi}. The ZMMI data tend to be highest
near the coasts, lowest in the Central United States, are relatively
symmetric, and have a mean of 55.05. The Hg ppb data tend to be highest
in the Northeastern United States, lowest elsewhere, are skewed, and
have a mean of 103.16 ppb. Also in Figure \ref{fig:zmmi} are separate
spatial semivariograms for ZMMI and Hg ppb. The spatial semivariogram
quantifies the halved average squared differences (semivariance) of
responses whose separation (distance) falls within a separation class.
The spatial semivariance is closely related to the spatial covariance,
and spatial semivariograms are often used to gauge the strength of
spatial dependence in data. Both ZMMI and Hg ppb seem to have moderately
strong spatial dependence (Figure \ref{fig:zmmi}), as the empirical
semivariance increases steadily with distance (meaning that observations
near one another tend to be more similar than observations far apart
from one another).

\begin{figure}
\centering
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/zmmi_map.jpeg}
  \caption{Spatial map of the ZMMI population.}
  \label{fig:zmmi_map}
\end{subfigure}
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/mercury_map.jpeg}
  \caption{Spatial map of the Hg ppb population.}
  \label{fig:mercury_map}
\end{subfigure} \\
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/zmmi_hist.jpeg}
  \caption{Histogram of the ZMMI population.}
  \label{fig:zmmi_hist}
\end{subfigure}
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/mercury_hist.jpeg}
  \caption{Histogram of the Hg ppb population.}
  \label{fig:mercury_hist}
\end{subfigure} \\
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/zmmi_sv.jpeg}
  \caption{Empirical semivariogram of the ZMMI population.}
  \label{fig:zmmi_sv_plot}
\end{subfigure}
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/mercury_sv.jpeg}
  \caption{Empirical semivariogram of the Hg ppb population.}
  \label{fig:mercury_sv_plot}
\end{subfigure}
\caption{Exploratory graphics representing populations for the zooplankton multi-metric indices (ZMMI) and mercury concentration in parts per billion (Hg ppb) in the 2012 National Lakes Assessment (NLA) data.}
\label{fig:zmmi}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width = 1\linewidth]{figures/data_coverage.jpeg}
  \caption{NLA data interval coverage for the four sampling-inference combinations. The columns indicate the response type. The solid black lines separate the sample sizes and the dashed black lines represent 95\% coverage.}
  \label{fig:data_figconf}
\end{figure}

@@

There are several benefits and drawbacks of the design-based and
model-based approaches for finite population spatial sampling and
inference. Some we have discussed, but others we have not, and they are
worthy of discussion. First, we discuss advantages of the design-based
approach. Design-based inference is often computationally efficient,
while model-based inference can be computationally burdensome,
especially for likelihood-based estimation methods like REML that rely
on the inverse of a covariance matrix. Design-based inference easily
handles binary data through a straightforward application of the
Horvitz-Thompson estimator. In contrast, analyzing binary data using
model-based inference generally requires a logistic mixed regression
model, the parameters of which can be difficult to estimate and
interpret (Bolker et al., 2009). An advantage of design-based inference
is that interval coverage is valid (has the proper coverage rate) as
long as 1) the sample is sufficiently large to ensure the statistic's
sampling distribution is approximately normal and 2) the variance
estimator is consistent (Brus and De Gruijter, 1997; Särndal et al.,
2003). This is because with the design-based approach, the sampling plan
and inclusion probabilities are specified directly by the researcher. An
advantage of SRS-DB not previously mentioned is that it is likely to be
valid given the consistency of its variance estimator (Särndal et al.,
2003). With the model-based approach, however, interval coverage is
unlikely to be valid if the model assumptions made do not not accurately
reflect reality. Whether model assumptions accurately reflect reality
can be a challenging and sometimes impossible question to answer
definitively.
